Density dependence

### 2020-10-22 analysis/paper/appendix_density_dependence.Rmd
Exclude saplings, include interaction.

source(here("R", "crossbasis.R"))

@@

@@

The exact coordinates (to the nearest 0.1m) of each plant relative to the plot were recorded for a subset of plots in three years. In 2006 and 2007, locations of plants were recorded to the nearest meter as part of a study to assess the effects of safe-site limitation on recruitment [@uriarte2010]. In 2009, plants were mapped to .... <!--# rubim study -->For each plot and year, we calculated the density of conspecific shoots in annuli at varying distances from focal plants, as in Teller et al. [-@teller2016]. Density was calculated for 5 annuli ranging from an inner radius of 0m and an outer radius of 0.5m (i.e. a circle of radius 0.5m) to an inner radius of 4.5m and and outer radius of 5m. If an annulus intersected with the edge of the plot, the density for that particular plant and annulus was considered as a missing value. Conspecific density was estimated as the number of H. acuminata shoots per m2 in each annulus. Number of shoots was used as a measure of plant size rather than height as it is less error prone to measure, is not affected by plants being knocked over, and is strongly correlated with plant height and leaf area in H. acuminata [@bruna2002; @brunaEffectsForestFragmentation2002].

@@

```{r data}
ha <- read_rds(here("analysis", "data", "derived_data", "ha_density.rds"))
ha$year %>% range()
```

```{r tidy}
ha <-
  ha %>%
  ungroup() %>% 
  arrange(ha_id_number, year) %>% 
  #scale year, and convert to factor since I'll be using as random effect
  mutate(year_post = as_factor(year - min(year)),
         log_shts = log(shts),
         log_shts_next = log(shts_next),
         habitat = as.factor(habitat),
         ha_id_number = as.factor(ha_id_number),
         plot = as.factor(plot))
```

@@

```{r}
g_growth <-
  gam(shts_next ~ 
        # s(ha_id_number, bs = "re") + #this doesn't seem to converge
        s(plot, bs = "re") +
        # s(log_shts, bs = "cr", by = habitat) + #prelim modeling shows this is a linear relationship (edf ~ 1)
        habitat * log_shts +
        s(Q, L,
          bs = "cb",
          # k = c(3,3),
          xt = list(bs = c("cr"))),
      family = gaussian(link = "log"),
      data = ha,
      method = "REML")
```

@@

Where $f \cdot w(x, d)$ is a two dimensional crossbasis smooth composed of two marginal functions: $f(x)$, the effect of conspecific density, $x$ , on growth, and the additional distance response function $w(d)$ that models the effect of density at different annuli, $d$, equal to the midpoint of the inner and outer radius of the 5 annuli for which density was calculated. Each dimension was modeled with 10 knots. With a penalized spline, the number of knots is not very important as ... <!--# finish thought and cite (maybe Colin Edwards?) --> Habitat is a parametric fixed effect with two levels (1-ha and continuous forest), $s(log(shts_t))$ is a penalized cubic spline describing the growth from year t to year t + 1, and a random effect of plot, nested within habitat. We did not include a random effect of plant ID, as the model did not converge. Although the response, shoot number in year t+1, is a discrete variable, a scaled t distribution with a log-link was used as it was a better fit than a poisson distribution ($\Delta$AIC = 810.5) and a log-normal distribution ($\Delta$AIC = 294).

```{r}
g_growth_pois <- update(g_growth, family = poisson(link = "log"))
g_growth_scat <- update(g_growth, family = scat(link = "log")) #because residuals are leptokurtic with gaussian
AICtab(g_growth, g_growth_pois, g_growth_scat)
```

@@

Number of shoots in the current year had a significant effect on size in year t + 1 (F = 353.8, edf = 5.2, P \< 0.0001). The relationship between $log(shoots_{t+1})$ and $log(shoots_t)$ was essentially linear for all but the largest size classes (Figure \@ref(fig:growth-plot)). There were no significant differences between habitat types (F = 0.012, df = 1, P = 0.912). The cross-basis smooth describing the effects of conspecific density on growth was significant (F = 2.78, edf = 18.4, P \< 0.001).

@@

Plants had a greater number of shoots in year t+1 as conspecific density increased, all else being average. The effects of conspecific density seemed strongest around 1.5 -- 2 meters from the focal plant (Figure \@ref(fig:growth-cb)).

```{r growth-slice, include=TRUE, fig.cap="this is a figure"}
pred_growth %>%
  filter(min_dist < 0.1) %>% 
  filter(lag == 1.75) %>% 
  ggplot(aes(x = x, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
  geom_jitter(data = ha %>%
                filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.))),
              aes(x = shts_1.5_2, y = shts_next),
              alpha = 0.2,
              inherit.aes = FALSE) +
  geom_line(color = "blue") + 
  geom_ribbon(alpha = 0.4) +

  labs(title = "annulus 1.5–2m", 
       x = "conspecific density (shoots/m^2)",
       y = "#shoots (t+1)") +
  theme_bw()
```

@@

Plant size had a significant, nearly quadratic, effect on survival (F = 126.59, edf = 2.1, P \< 0.001) (Figure \@ref(fig:surv-plot)). However, there were no significant effects of habitat ($\chi^2$ = 0.26, df = 1, P = 0.613) or conspecific density ($\chi^2$ = 126.59, edf = 2, P = 0.074) on survival.

@@

@@

ggplot(fitted, aes(x = exp(log_shts), y = plogis(.fitted))) +
  geom_line(color = "blue") +
  geom_ribbon(aes(ymin = plogis(.fitted - .se.fit*1.96), ymax = plogis(.fitted + .se.fit*1.96)), alpha = 0.4) +
  geom_rug(data = rugdat,
           aes(x = survived),
           inherit.aes = FALSE,
           sides = "t",
           alpha = 0.2) +
  geom_rug(data = rugdat,
           aes(x = died),
           inherit.aes = FALSE,
           sides = "b",
           alpha = 0.2) +
  stat_summary_bin(geom = "point",
                   fun = mean,
                   aes(y = surv),
                   bins = 1000,
                   data = ha %>%
                     filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.)))) +
  labs(x = "#shoots (t)", y = "survival probability") +
  scale_x_continuous(n.breaks = 7) +
  theme_bw()
```

@@

```{r eval=FALSE, include=FALSE}
pred_growth %>%
  filter(min_dist < 0.1) %>% 
  ungroup() %>% 
  filter(fitted == max(fitted))

pred_growth %>%
  filter(min_dist < 0.1) %>% 
  filter(near(x, 5.01, tol = 0.01)) %>% 
  ggplot(aes(x = lag, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
  # geom_jitter(data = ha %>%
  #               filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.))),
  #             aes(x = shts_1.5_2, y = shts_next),
  #             alpha = 0.2,
  #             inherit.aes = FALSE) +
  # geom_smooth(color = "blue", se = FALSE) + 
  geom_line(color = "blue") +
  geom_ribbon(alpha = 0.4) +

@@

```{r}
mnorm <-
  glmer(
    shts_next ~ 
      log_shts *
      habitat *
      shts_0_3 +
      (1|ha_id_number) +
      (1|plot) +
      (1|year_post),
    family = gaussian(link = "log"),
    data = ha
  )
# Random effects
mnorm2 <- update(mnorm, . ~ . - (1|ha_id_number))
AICtab(mnorm, mnorm2) #keep ID number
mnorm3 <- update(mnorm, .~. - (1|plot)) 
AICtab(mnorm, mnorm2, mnorm3) #remove plot
mnorm4 <- update(mnorm3, .~.-(1|year_post)) #remove year
AICtab(mnorm, mnorm2, mnorm3, mnorm4)

>>>>

source(here("R", "crossbasis.R"))
source(here("R", "utils.R"))

@@

-   Remove saplings and update stats

@@

The exact coordinates (to the nearest 0.1m) of each plant relative to the plot were recorded for a subset of plots in three years. In 2006 and 2007, locations of plants were recorded to the nearest meter as part of a study to assess the effects of safe-site limitation on recruitment [@uriarte2010]. In 2009, plants were mapped to .... <!--# rubim study -->For each plot and year, we calculated the density of conspecific shoots in annuli at varying distances from focal plants, as in Teller et al. [-@teller2016]. Density was calculated for 5 annuli ranging from an inner radius of 0m and an outer radius of 0.5m (i.e. a circle of radius 0.5m) to an inner radius of 4.5m and and outer radius of 5m. If an annulus intersected with the edge of the plot, the density for that particular plant and annulus was considered as a missing value. Conspecific density was estimated as the size of H. acuminata shoots per m2 in each annulus. <!--# Add analyses using height and shoots + height -->Number of shoots was used as a measure of plant size rather than height as it is less error prone to measure, is not affected by plants being knocked over, and is strongly correlated with plant height and leaf area in H. acuminata [@bruna2002; @brunaEffectsForestFragmentation2002].

@@

```{r data}
ha <- read_rds(here("analysis", "data", "derived_data", "ha_density.rds"))
ha$year %>% range()
ha$code_notes %>% unique()
```

```{r tidy}
ha <-
  ha %>%
  filter(code_notes != "sdlg (1)" | is.na(code_notes)) %>% #remove seedlings
  arrange(ha_id_number, year) %>% 
  #scale year, and convert to factor since I'll be using as random effect
  mutate(year_post = as_factor(year - min(year)),
         log_shts = log(shts),
         log_shts_next = log(shts_next),
         # log_ht = log(ht),
         # log_ht_next = log(ht_next),
         habitat = as.factor(habitat),
         ha_id_number = as.factor(ha_id_number),
         plot = as.factor(plot))
```

@@

```{r}
g_growth_scat <-
  gam(shts_next ~ 
        # s(ha_id_number, bs = "re") + #this doesn't seem to converge
        s(plot, bs = "re") +
        # s(log_shts, bs = "cr", by = habitat) + #prelim modeling shows this is a linear relationship (edf ~ 1)
        log_shts *
        habitat +
        s(Q, L,
          bs = "cb",
          # k = c(3,3),
          xt = list(bs = c("cr"))),
      family = scat(link = "log"),
      data = ha,
      method = "REML")
```

@@

Where $f \cdot w(x, d)$ is a two dimensional crossbasis smooth composed of two marginal functions: $f(x)$, the effect of conspecific density, $x$ , on growth, and the additional distance response function $w(d)$ that models the effect of density at different annuli, $d$, equal to the midpoint of the inner and outer radius of the 5 annuli for which density was calculated. Each dimension was modeled with 10 knots. With a penalized spline, the number of knots is not very important as ... <!--# finish thought and cite (maybe Noam Ross) --> Habitat is a parametric fixed effect with two levels (1-ha and continuous forest), $s(log(shts_t))$ is a penalized cubic spline describing the growth from year t to year t + 1, and a random effect of plot, nested within habitat. We did not include a random effect of plant ID, as the model did not converge. Although the response, shoot number in year t+1, is a discrete variable, a scaled t distribution with a log-link was used as it was a better fit than a poisson distribution ($\Delta$AIC = 810.5) and a log-normal distribution ($\Delta$AIC = 294).

```{r eval=FALSE, include=FALSE}
g_growth_pois <- update(g_growth, family = poisson(link = "log"))
g_growth_norm <- update(g_growth, family = gaussian(link = "log")) #residuals are leptokurtic with gaussian
AICtab(g_growth, g_growth_pois, g_growth_scat)
```

@@

Number of shoots in the current year had a significant effect on size in year t + 1 ($\chi^2$ = 796.8, df = 1, P \< 0.0001) (Figure \@ref(fig:growth-plot)). There was no significant effect of habitat or the interaction between habitat and size on size in year t + 1. The cross-basis smooth describing the effects of conspecific density on growth was significant ($\chi^2$ = 23.2, edf = 6.4, P = 0.004).

@@

Plants had a greater number of shoots in year t+1 as conspecific density increased, all else being average. The effects of conspecific density were strongest around 1.5 -- 2 meters from the focal plant (Figure \@ref(fig:growth-cb)).

```{r growth-slice, include=TRUE, fig.cap="this is a figure"}
pred_growth %>%
  filter(min_dist < 0.1) %>%
  filter(nearest(lag, 2)) %>% 
  ggplot(aes(x = x, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
  geom_jitter(data = ha %>%
                filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.))),
              aes(x = shts_1.5_2, y = shts_next),
              alpha = 0.2,
              inherit.aes = FALSE) +
  geom_line(color = "blue") + 
  geom_ribbon(alpha = 0.4) +

  labs(title = "Near 2m from focal plant", 
       x = "conspecific density (shoots/m^2)",
       y = "#shoots (t+1)") +
  theme_bw()
```

@@

Plant size had a significant, nearly quadratic (on a logit scale), effect on survival in continuous forest ($\chi^2$ = 112.2, edf = 2.1, P \< 0.001), but a linear (on a logit scale) effect in 1 ha fragments ($\chi^2$ = 15.63, edf = 1, P \< 0.001) (Figure \@ref(fig:surv-plot)). However, there were no significant effects of habitat ($\chi^2$ = 0.002, df = 1, P = 0.964) or conspecific density ($\chi^2$ = 5.326, edf = 2, P = 0.0699) on survival.

@@

#TODO: make this have both 1-ha and CF

newdata <-
  tibble(
    log_shts = rep(seq(0, 2.565, length.out = 100), 2),
    habitat = rep(c("1-ha", "CF"), each = 100),
    plot = "new",
    Q = Q_new,
    L = L_new
  )

@@

ggplot(fitted, aes(x = exp(log_shts), y = plogis(.fitted), color = habitat)) +
  geom_line() +
  geom_ribbon(aes(
    ymin = plogis(.fitted - .se.fit * 1.96),
    ymax = plogis(.fitted + .se.fit * 1.96),
    color = NULL,
    fill = habitat
  ), alpha = 0.4)+
  geom_rug(data = rugdat,
           aes(x = survived, color = habitat),
           inherit.aes = FALSE,
           sides = "t",
           alpha = 0.2) +
  geom_rug(data = rugdat,
           aes(x = died, color = habitat),
           inherit.aes = FALSE,
           sides = "b",
           alpha = 0.2) +
  stat_summary_bin(geom = "point",
                   fun = mean,
                   aes(y = surv),
                   bins = 1000,
                   data = ha %>%
                     filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.)))) +
  labs(x = "#shoots (t)", y = "survival probability") +
  scale_x_continuous(n.breaks = 7) +
  theme_bw()
```

@@

```{r eval=FALSE, include=FALSE}
pred_growth %>%
  filter(min_dist < 0.1) %>% 
  filter(nearest(x, 5)) %>% 
  ggplot(aes(x = lag, y = exp(fitted), ymin = exp(fitted - 1.96*se.fit), ymax = exp(fitted + 1.96*se.fit))) +
  # geom_jitter(data = ha %>%
  #               filter(across(shts_0.5_1:shts_4.5_5, ~!is.na(.))),
  #             aes(x = shts_1.5_2, y = shts_next),
  #             alpha = 0.2,
  #             inherit.aes = FALSE) +
  # geom_smooth(color = "blue", se = FALSE) + 
  geom_line(color = "blue") +
  geom_ribbon(alpha = 0.4) +

@@

```{r}
mnorm <-
  glmer(
    shts_next ~ 
      log_shts *
      habitat *
      shts_0_3 +
      (1|ha_id_number) +
      (1|plot) +
      (1|year_post),
    family = gaussian(link = "log"), #can't use scat() in glmer()
    data = ha
  )
# Random effects
mnorm2 <- update(mnorm, . ~ . - (1|ha_id_number))
AICtab(mnorm, mnorm2) #keep ID number
mnorm3 <- update(mnorm, .~. - (1|plot)) 
AICtab(mnorm, mnorm2, mnorm3) #remove plot
mnorm4 <- update(mnorm3, .~.-(1|year_post)) #remove year
AICtab(mnorm, mnorm2, mnorm3, mnorm4)
